\chapter*{}
\thispagestyle{empty}
\addcontentsline{toc}{chapter}{Acknowledgment}

\begin{center}
    \vspace*{-0.5cm}
    {\Large \textbf{ACKNOWLEDGMENT}}
\end{center}

\vspace{0.6cm}

We would like to express our sincere gratitude to all those who have contributed to the successful completion of this major project.

First and foremost, we are deeply grateful to our project supervisor, \textbf{Dr.~Ramesh Chandra Mishra}, Assistant Professor, Department of Electronics and Communication Engineering, IIIT Manipur, for his invaluable guidance, continuous encouragement, and insightful suggestions throughout the course of this project. His expertise in wireless communications and deep learning has been instrumental in shaping this work.

We extend our heartfelt thanks to \textbf{Dr.~Nagesh Ch}, Head of Department of Electronics and Communication Engineering, for providing the necessary support, facilities and resources that enabled us to undertake this research.

We are especially thankful to \textbf{Dr.~Chittotosh Ganguly} for his initial guidance and help, which set the foundation for this work. We also thank all other faculty members of the Department of Electronics and Communication Engineering for their support and for sharing their knowledge throughout this project.

We would also like to acknowledge the authors of the research papers and technical reports cited in this work, whose pioneering contributions to NOMA systems and Graph Neural Networks have inspired and guided this research.

Our special thanks to our families and friends for their unwavering support, patience, and motivation throughout this journey. Their encouragement has been a constant source of strength.

Finally, we bow in reverence to the divine grace of God for blessing us with the strength, wisdom, and perseverance to complete this project successfully.

\vspace{2cm}

\noindent
\begin{minipage}[t]{0.48\textwidth}
\raggedright
\textbf{Anisha Dwivedi}\\
Roll No.: 220102020
\end{minipage}
\hfill
\begin{minipage}[t]{0.48\textwidth}
\raggedright
\textbf{Shailendra Shukla}\\
Roll No.: 220102009
\end{minipage}

\newpage
