\chapter*{Acknowledgment}
\addcontentsline{toc}{chapter}{Acknowledgment}

\vspace{1cm}

I would like to express my sincere gratitude to all those who have contributed to the successful completion of this major project.

First and foremost, I am deeply grateful to my project supervisor, \textbf{[Supervisor Name]}, [Designation], Department of Computer Science and Engineering, IIIT Manipur, for their invaluable guidance, continuous encouragement, and insightful suggestions throughout the course of this project. Their expertise in wireless communications and deep learning has been instrumental in shaping this work.

I extend my heartfelt thanks to \textbf{[Head of Department Name]}, Head of the Department of Computer Science and Engineering, for providing the necessary facilities and resources that enabled me to undertake this research.

I am thankful to all the faculty members of the Department of Computer Science and Engineering for their support and for sharing their knowledge, which laid the foundation for this project.

I would also like to acknowledge the authors of the research papers and technical reports cited in this work, whose pioneering contributions to NOMA systems and Graph Neural Networks have inspired and guided this research.

My special thanks to my family and friends for their unwavering support, patience, and motivation throughout this journey. Their encouragement has been a constant source of strength.

Finally, I am grateful to the Almighty for giving me the strength, wisdom, and perseverance to complete this project successfully.

\vspace{2cm}

\begin{flushright}
\textbf{[Your Name]} \\
Roll No: [Your Roll Number]
\end{flushright}

\newpage
