\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

\vspace{1cm}

I, \textbf{Anisha Dwivedi and Shailendra Shukla}, Roll No: \textbf{[Your Roll Numbers]}, students of B.Tech in Computer Science and Engineering at Indian Institute of Information Technology, Manipur, hereby declare that the project work entitled \textbf{"Deep Learning for User Pairing and Power Allocation in Power-Domain NOMA Systems"} submitted to the Department of Computer Science and Engineering, IIIT Manipur, is a record of an original work done by us under the guidance of \textbf{Dr. Ramesh Ch. Mishra}, Assistant Professor, Department of Computer Science and Engineering, IIIT Manipur.

\vspace{1cm}

I further declare that:
\begin{enumerate}
    \item The work reported in this project has been performed by me.
    \item I have not submitted this work elsewhere for any other degree or diploma.
    \item Whenever contributions of others are involved, every effort is made to indicate this clearly with due acknowledgment and reference to the literature.
    \item All sources of information have been specifically acknowledged by means of proper references.
    \item The project does not contain any plagiarized material beyond the permissible limit.
\end{enumerate}

\vspace{2cm}

\begin{flushright}
\textbf{Anisha Dwivedi} \\
Roll No: [Your Roll Number] \\
Date: \_\_\_\_\_\_\_\_\_\_\_\_ \\
Place: Imphal, Manipur \\[1cm]

\textbf{Shailendra Shukla} \\
Roll No: [Your Roll Number] \\
Date: \_\_\_\_\_\_\_\_\_\_\_\_ \\
Place: Imphal, Manipur
\end{flushright}

\newpage
