\chapter*{Acknowledgment}
\addcontentsline{toc}{chapter}{Acknowledgment}

\vspace{1cm}

I would like to express my sincere gratitude to all those who have contributed to the successful completion of this major project.

First and foremost, we are deeply grateful to our project supervisor, \textbf{Dr. Ramesh Ch. Mishra}, Assistant Professor, Department of Computer Science and Engineering, IIIT Manipur, for his invaluable guidance, continuous encouragement, and insightful suggestions throughout the course of this project. His expertise in wireless communications and deep learning has been instrumental in shaping this work.

We extend our heartfelt thanks to \textbf{[Head of Department Name]}, Professor and Head of the Department of Computer Science and Engineering, for providing the necessary facilities and resources that enabled us to undertake this research.

We are thankful to all the faculty members of the Department of Computer Science and Engineering for their support and for sharing their knowledge, which laid the foundation for this project.

We would also like to acknowledge the authors of the research papers and technical reports cited in this work, whose pioneering contributions to NOMA systems and Graph Neural Networks have inspired and guided this research.

Our special thanks to our family and friends for their unwavering support, patience, and motivation throughout this journey. Their encouragement has been a constant source of strength.

Finally, we are grateful to the Almighty for giving us the strength, wisdom, and perseverance to complete this project successfully.

\vspace{2cm}

\begin{flushright}
\textbf{Anisha Dwivedi} \\
Roll No: [Your Roll Number] \\[1cm]
\textbf{Shailendra Shukla} \\
Roll No: [Your Roll Number]
\end{flushright}

\newpage
